Spatiotemporal Multi-Resolution Approximation of the Amari Type Neural Field Model

### 2012-02-02 Multiresolution_framework/ltx/manuscript.tex
spell checking

\begin{abstract}
%\boldmath
We develop a multi-resolution approximation (MRA) framework for the integro-difference equation (IDE) neural field model based on semi-orthogonal cardinal B-spline wavelets. State and parameter estimation is performed using the Expectation Maximization (EM) algorithm. A synthetic example is provided
to demonstrate the framework.
\end{abstract} 
\begin{keyword}
  Neural Field Model, multi-resolution approximation (MRA), Expectation Maximisation (EM) algorithm
\end{keyword}

@@

The first work (to the authors best knowledge) describing data-driven mesoscopic neural modelling used a neural mass model to fit EEG data~\citep{Valdes1999}. This approach was extended to coupled neural masses through a Bayesian estimation scheme dubbed Dynamic Causal Modelling (DCM)~\citep{David2003}. Following this work, data-driven modelling was extended to continuum field equations that explained the richer dynamics of spatiotemporal neural fields \citep{Galka2008,schiff2008kalman,Daunizeau2009}. Most recently, a framework was developed where a finite element model (FEM) of the neural field (via a global Galerkin projection) was formed, using a basis function decomposition, to transform the PDE neural field equations into a finite dimension system to facilitate efficient state and parameter estimation~\citep{Freestone2011}. This paper is an extension to the aforementioned study, where we derive a neural field state-space model that accounts for the multi-resolution architecture and spatial dynamics of the human cortex. In this way, a flexible framework is created, whereby both macroscopic and microscopic behaviour of the system can be represented simultaneously and completely.

@@

The multi-resolution structure has been described by numerous methods. For example, this idea of multi-resolution can be thought of a generalisation of modelling the layers of the cortex separately, where the connectivity kernel, or neural foot-print, has varying widths \citep{Wilson1973}.

@@

The term $e_t(\mathbf r)$ is a zero mean Gaussian disturbance, spatially coloured but temporally white, with covariance function 
\begin{equation}
cov\left(e_{t}\left(\mathbf{r}\right),e_{t+t'}\left(\mathbf{r'}\right)\right)=\eta(\mathbf{r}-\mathbf{r'})\delta(t-t'),
\label{eq:FieldDisturbance}
\end{equation}
where $\delta(\cdot)$ is the Dirac-delta function. %$ = \varsigma v_t(\mathbf{r})$ \cite{VanRotterdam1982,Murphy2009}.

@@

The goal of this section is to demonstrate how to formulate a state-space model in a canonical form such that standard algorithms may be applied to estimate the multi-resolution connectivity structure. Therefore, the goal is to form a parameterised equation for coefficients of the field basis functions $\mathbf{x}_t$, which becomes the state vector.

@@

\section{B-Spline MRAIDE in State-Space }
\subsection{A Brief Introduction to B-Spline Wavelets}
B-spline functions are one of the most basic elements for wavelet construction. These functions are composed of piecewise polynomials combined smoothly at some breaking points (knots), which are equally spaced in the case of cardinal B-splines. The degree of smoothness is determined by the order of splines \citep{Goswami1999}, where the higher the order the smoother the resulting B-spline becomes (see Figure~\ref{fig:Figure0}).
\begin{figure}[!t]
\centering
\includegraphics{./Graph/Figure0.pdf}
% \includegraphics{./Graph/Figure0b.pdf}
\caption{{\bf Examples of cardinal B-spline functions}. Piecewise polynomial functions (coloured lines) are joined at the break points (black dots). (\textbf{A}) B-spline function of order 2 (linear B-spline). (\textbf{B}) B-spline function of order 8.}
\label{fig:Figure0}
\end{figure}

B-spline wavelet and scaling functions were formulated independently by \citet{Chui1992b,Chui1992,Unser1993}.  The scaling functions are $m$-th order B-splines and the compact support wavelet functions are a linear combination of scaling functions. B-spline wavelet and scaling functions approach optimal time/frequency localisation as the order of the spline increases. In fact, for the cubic B-spline scaling and wavelet functions it is already close to the optimal limit for Gaussian functions and they have a better approximation rate than other wavelets with the same number of vanishing moments \citep{Unser1999}.

B-spline wavelet and scaling functions are particularly suited in this framework due to the property of being able to analytically define the convolution and inner product to form MRAIDE components. The following summarises some important characteristics of B-splines relevant to the proposed modelling framework.

@@

By exploiting \eqref{eq:BsplineConvolution} and \eqref{eq:BsplineInnerProduct}, the integrals in \eqref{eq:Lambdax}, \eqref{eq:Lambdatheta} and \eqref{eq:CovMatrix} can be computed analytically. In constructing $\boldsymbol\Lambda_{x}$, it is important to note that B-spline scaling and wavelet functions possess the following orthogonality properties \citep{Unser1993}: 
\begin{equation}
  \left\langle \varphi_{m;j_1,l_1}(r),\varphi_{m;j_2,l_2}(r)\right\rangle =0  \quad \mathrm{for} \quad j_1\neq j_2
 \label{eq:MRA-PsiPsiOrthogonality} 
 \end{equation}
 \begin{equation}
  \left\langle N_{m;j_1,l_1}(r),\varphi_{m;j_2,l_2}(r)\right\rangle =0  \quad \mathrm{for} \quad j_1\leq j_2.
 \label{eq:MRA-PhiPsiOrthogonality}
 \end{equation}
  \parham{Note from \eqref{eq:MRA-PsiPsiOrthogonality} B-spline wavelets are semiorthogonal as they are not orthogonal with respect to translation on a given scale.}

@@

To demonstrate the MRAIDE framework we shall consider the parameters to be static and we shall also assume that the relationship between the mean membrane potential and the mean firing rate is predominantly linear. Note that there is a difference between forward and inverse modelling when making these assumptions. For example, by using a linear activation function in the estimator we are not assuming that there is necessarily a linear relationship between the membrane voltage and firing rate, but that the statistics of the neural masses fall within a linear operating region over the estimation time period.

@@

By assuming a linear activation function of the form
\begin{equation}
	f(v_t(\mathbf{r})) = \varsigma v_t(\mathbf{r})
\end{equation}
the state transition equation~\eqref{eq:ApproxDiscreteTimeModel4} simplifies to
\begin{equation}\label{eq:ApproxDiscreteTimeModel_Linear}
	\mathbf{x}_{t+1} = 
	\xi \mathbf{x}_t + 
	\varsigma T_s \mathbf{\Lambda}_{x}^{-1} \int_{\Omega}\boldsymbol\mu\left(\mathbf{r}\right)\int_\Omega { 
	    \boldsymbol\theta^\top\boldsymbol\lambda\left(\mathbf{r}-\mathbf{r}'\right)
	    \boldsymbol\mu^\top\left(\mathbf{r}'\right) 
	\, \mathrm{d}\mathbf{r}'\mathrm{d}\mathbf{r}} \mathbf{x}_t
	+ \mathbf{w}_t.
\end{equation}
Now we can further simplify the system by defining
\begin{align}
	\label{eq:Lambdatheta}
	 \mathbf{\Lambda}_{\theta} &\triangleq \int_{\Omega}\boldsymbol\mu\left(r\right) \int_\Omega { 
		   \boldsymbol\theta^\top\boldsymbol\lambda\left(r-r'\right)
		    \boldsymbol\mu^\top\left(r\right)\ \mathrm{d}r'\mathrm{d}r}
\end{align}
to give the state transition equation
\begin{align}\label{eq:StateEquation}
 \mathbf x_{t+1} &=\mathbf A(\boldsymbol \theta) \mathbf x_t+ \mathbf w_t\\
\label{eq:A_theta}
 \mathbf A(\boldsymbol \theta) &= T_s\varsigma\mathbf{\Lambda}_{x}^{-1}\mathbf{\Lambda}_{\theta}+\xi\mathbf I.
\end{align} 
The problem of joint state and parameter estimation for the state-space model given by \eqref{eq:StateEquation} and \eqref{eq:ReducedObservationEquation} can be formulated as 
\begin{equation}
	{\hat{\mathbf X},\hat{\boldsymbol\theta}}=\arg\max_{\mathbf X,\boldsymbol\theta}p(\mathbf Y;\boldsymbol\theta),
 \end{equation}  
where $\mathbf X$ and $\mathbf Y$ denote the entire sequences of the states and observations respectively, $\boldsymbol\theta$ is the parameter set and $p(\mathbf Y;\boldsymbol\theta)$ denotes the probability density function of the observations parameterised  by $\boldsymbol\theta$, known as the likelihood function. This problem cannot be solved separately as for the state estimation the knowledge of the system's parameters is required and for the parameter estimation the states of the system need to be known. One approach is to iteratively estimate the states and parameters of the system and monitor the convergence of the algorithm. A natural framework to perform this is the well known Expectation Maximisation (EM) algorithm \cite{Dempster1977,Shumway2000} to infer both the parameter set and the states from the observations corresponding to the connectivity kernel, the neural field and electrophysiological data in our application. There are two main applications of the EM algorithm:  first is when the observations are incomplete due to a faulty observation process or existing limitations and the second  when the direct maximisation of the likelihood function is difficult but can be simplified by assuming the existence of an additional however missing (or hidden) variables \cite{Bilmes1998}, the weights of the field basis functions in this case. The EM algorithm, when used in the latter context \cite{Dewar2009}, yields the maximum likelihood kernel estimate, i.e.
\begin{equation}
	\boldsymbol\theta_{\text{ML}}=\arg\max_{\boldsymbol\theta}~p(\mathbf Y;\boldsymbol\theta),
 \end{equation}   
and the posterior distribution of the field over time.

@@

In summary, to solve the joint state and parameter estimation for linear state-space models one can adapt the EM algorithm, solving the smoothing problem for the E-step and constructing and maximising the lower bound in the M-step. A schematic representation of the EM algorithm is represented in Figure~\ref{fig:EMBlockDiagram}.  

@@

\caption{{\bf Schematic of the EM algorithm}. The RTS smoother (E-step) and the maximisation of the lower bound, $\mathcal Q(\boldsymbol \theta,\boldsymbol\theta')$, (M-step) are shown. The dashed box shows the iterative two step algorithm once it is initialised.}

@@

  As mentioned earlier the E-step for state-space models is equivalent to the smoothing problem. We use the RTS smoother in order to infer the states of the system. The RTS smoother uses the Kalman filter \cite{Kalman1960} in forward pass along withe a backward iteration to calculate its outputs: smoothed states $\hat{\mathbf x}^b_t$ and covariances, $\mathbf P^b_t=\mathrm{cov}(\mathbf{x}_t)$. The RTS  algorithm is given in Table~\ref{alg:MRA-RTS}. Following \cite{Gibsona2005}, an additional backwards recursion is also included (Step~5 of the RTS algorithm)	to calculate cross-covariance matrix, $\mathbf M_t=\mathrm{cov}(\mathbf{x}_{t},\mathbf{x}_{t+1})$, which is also required in forming the lower bound to be maximised in the M-step. Note that the following quantities from the forward iteration (filtering) should be stored when implementing the RTS smoother: predicted and corrected estimates for both state estimates ($\mathbf{x}_t^{f-}$, $\mathbf{x}_t^{f}$) and covariance matrices ($\mathbf P_t^{f-}$,$\mathbf P_t^f$). Also the final value of the Kalman gain, $\mathcal K_T$, should be stored in order to initialise the backward cross-covariance computation. 

@@

\caption{\textbf{Algorithm for the RTS smoother} This table shows the steps in the  Rauch-Tung-Striebel smoother algorithm. The lower bound is maximised after each iteration to update the parameter estimates.}

@@

where the constant term can be ignored as it is independent of the parameters. Under the condition that the state distribution is Gaussian --- ignoring the normalising term, $1/(2\pi)^{\frac{n_x}{2}}\mid\boldsymbol\Sigma_w\mid^{-\frac{1}{2}}$ --- the conditional distribution $p(\mathbf x_{t+1} | \mathbf x_{t};\boldsymbol\theta)$ can be written as

@@

Rearranging and taking the expectation of the log-likelihood function over all time instants gives the required lower-bound which is to be maximised for the optimal parameter estimates. Note that the expectation distributes over the trace sum. Therefore, 

@@

\subsection{Stopping rule}
The stopping criterion is usually associated with either the change in the parameter estimates or the log-likelihood variation \cite{McLachlan1997}. The stopping rule adopted herein is
\begin{equation}
 \left(\parallel \mathbf{A} \parallel_{F}^{(i)}-\parallel \mathbf{A} \parallel_{F}^{(i-1)}\right)<\epsilon,
 \end{equation}
 where $\epsilon$ is a threshold value and $\parallel \mathbf{A} \parallel_{F}^{(i)}$ and $ \parallel \mathbf{A} \parallel_{F}^{(i-1)}$ are the Frobenius norms  of the successive estimates of $\mathbf{A} $ matrices which is defined by \citet{Meyer2000}
 \begin{equation}
  \parallel \mathbf{A} \parallel_{F}=\sqrt{\sum_{i,j=1}^{n_x}\mid a_{i,j} \mid^2}=\sqrt{\mathrm{tr} (\mathbf A^{\top}\mathbf A)}.
 \end{equation}
\subsection{Computational complexity} 
All $n_x^2$ blocks of $\mathbf U$, $\boldsymbol\Gamma_1$ and $\boldsymbol\Gamma_2$  can be computed as a one-off before the commencement of the EM iterations, which increases the speed of the M-step significantly compared to the implementation in \citet{Dewar2009}. The complexity of each step of the algorithm is summarised in Table~\ref{table:MRA-ComputationalComplexity} where for ease of comparison the computational complexity of the method in \citet{Dewar2009} is also identified. It can be seen from the table that the state dimension is an important factor at each iteration of the estimation algorithm, emphasizing the advantage of the proposed algorithm compared to \citet{Dewar2009}. This also suggests the use of an efficient algorithm to choose the minimum number of basis functions (states of the system) while achieving the desired details of the underlying field.  
\begin {table}
\begin{center}
\scalebox{1}{\begin{tabular}{lllc}
\hline \hline
Variable &Equation&Order&Order obtained from \citet{Dewar2009}\\ \hline\\
$\Xi_0, \Xi_1$&\eqref{eq:MRA-Xi0}, \eqref{eq:MRA-Xi1}&$O(Tn_x^2)$& -\\
$\boldsymbol\upsilon_0, \boldsymbol\upsilon_1$&\eqref{eq:epsilon0},\eqref{eq:epsilon1} &$O(n_x^2n_{\theta})$ &$O(n_x^2n_{\theta})+O(n_x^3)$ \\
$\boldsymbol\Upsilon$&\eqref{eq:Epsilon}&$O(n_x^2n_{\theta}^2)$&$O(n_x^4n_{\theta}^2)$\\
$\mathbf A(\boldsymbol\theta)$&\eqref{eq:A_theta}&$O(n_x^3)$&-\\
$\boldsymbol\Lambda_{\theta}$&\eqref{eq:Lambdatheta}&$O(n_x^2n_{\theta})$&-\\
$\hat{\boldsymbol\theta} $&\eqref{eq:MRA-thetahat}&$O(n_{\theta}^3)$&-\\
\hline \hline
\end{tabular}}
 \caption {{\bf The Computational complexity of the estimation algorithm}. A comparison between the algorithm proposed herein and that of \citet{Dewar2009}.} 
\label{table:MRA-ComputationalComplexity}
\end{center}
\end {table}

@@

Two different experiments were designed to test the framework. In the first one, 200 realisations of 1 second of data were used, where the field disturbance, $e_t$, and the observation noise, $\boldsymbol\varepsilon_t$, were regenerated each run. This way the net shape of the connectivity kernel could be formed by averaging over the parameters estimates from each realisation. In the second experiment, a single data set was generated and the field estimations are compared for five models, each accounting for a different level of approximation, i.e. $j=0$ to $j=4$.

For each of the experiments, the estimation was applied to the final 900~ms, allowing the forward model's dynamics to stabilise from the initial conditions. The initial state, parameters and the support of the connectivity kernel were unknown to the estimator. The observation noise was set to $\boldsymbol\Sigma_{\varepsilon}=0.1 \times \mathbf{I}_{n_y}$ and the disturbance covariance function was set to $\gamma(r-r') = \varphi_{3,-2}(r-r')$ (B-spline, $j=3$).  The firing rate slope, $\varsigma$ was set to $0.56~mV^{-1}$ \citep{Wendling2005}. The observation kernel was modelled using a B-spline function, with a width of 0.8~mm at half the maximum amplitude where the distance between adjacent sensors was $0.5$~mm, resulting into $n_y = 161$ observations. The spacing and bandwidth of the sensors allowed the full spatial bandwidth of the field to be observed.  With this number of observations, the estimation problem was well-posed, i.e. the number of states, $n_x < n_y$ at each level of the decomposition. This is because the scaling functions are orthogonal to the wavelets at the same and higher levels of decomposition, and  the wavelets are orthogonal to each other across different scales (see (\ref{eq:MRA-PsiPsiOrthogonality}-\ref{eq:MRA-PhiPsiOrthogonality})).

@@

The actual connectivity kernel and the decomposition is plotted in Figure~\ref{fig:KernelEstimate}(a). No assumptions where made about the shape of the kernel. The reconstructed kernel is in good accordance with the actual kernel, where the actual kernel lies inside the confidence interval. The large standard deviation is due to the high number of parameters need to be estimated. The small error in the estimate is attributed to the MRA of the system used to form the estimator. Figure~\ref{fig:KernelEstimate}(b) and (c) illustrate the kernel scaling and wavelet functions, respectively. A total number of 25 basis functions was used to reconstruct the kernel, comprising 13 and 12 scaling and wavelet functions at $j=1$ accordingly. 
\begin{figure}[!t]
\includegraphics{./Graph/Figure5.pdf}
\includegraphics{./Graph/Figure6.pdf}
\caption{ {\bf Connectivity kernel estimate}. (\textbf{a}) The actual kernel and its components are shown by solid and dashed black lines, respectively. The mean kernel estimates over 200 realisations and confidence interval are shown by the red line and red shaded region ($\pm2$ std.). (\textbf{b}) Kernel scaling and wavelets functions, respectively, at $j=1$ used in the estimation algorithm.}
\label{fig:KernelEstimate}
\end{figure}

>>>>

\begin{abstract}
%\boldmath
We develop a multi-resolution approximation (MRA) framework for the integro-difference equation (IDE) neural field model based on semi-orthogonal cardinal B-spline wavelets. State and parameter estimation is performed using the Expectation Maximization (EM) algorithm. A synthetic example is provided
to demonstrate the framework.
\end{abstract} 
\begin{keyword}
  Neural Field Model, multi-resolution approximation (MRA), Expectation Maximization (EM) algorithm
\end{keyword}

@@

The first work (to the authors best knowledge) describing data-driven mesoscopic neural modeling used a neural mass model to fit EEG data~\citep{Valdes1999}. This approach was extended to coupled neural masses through a Bayesian estimation scheme dubbed Dynamic Causal Modeling (DCM)~\citep{David2003}. Following this work, data-driven modeling was extended to continuum field equations that explained the richer dynamics of spatiotemporal neural fields \citep{Galka2008,schiff2008kalman,Daunizeau2009}. Most recently, a framework was developed where a finite element model (FEM) of the neural field (via a global Galerkin projection) was formed, using a basis function decomposition, to transform the PDE neural field equations into a finite dimension system to facilitate efficient state and parameter estimation~\citep{Freestone2011}. This paper is an extension to the aforementioned study, where we derive a neural field state-space model that accounts for the multi-resolution architecture and spatial dynamics of the human cortex. In this way, a flexible framework is created, whereby both macroscopic and microscopic behavior of the system can be represented simultaneously and completely.

@@

The multi-resolution structure has been described by numerous methods. For example, this idea of multi-resolution can be thought of a generalization of modeling the layers of the cortex separately, where the connectivity kernel, or neural foot-print, has varying widths \citep{Wilson1973}.

@@

The term $e_t(\mathbf r)$ is a zero mean Gaussian disturbance, spatially colored but temporally white, with covariance function 
\begin{equation}
cov\left(e_{t}\left(\mathbf{r}\right),e_{t+t'}\left(\mathbf{r'}\right)\right)=\eta(\mathbf{r}-\mathbf{r'})\delta(t-t'),
\label{eq:FieldDisturbance}
\end{equation}
where $\delta(\cdot)$ is the Dirac-delta function. %$ = \varsigma v_t(\mathbf{r})$ \cite{VanRotterdam1982,Murphy2009}.

@@

The goal of this section is to demonstrate how to formulate a state-space model in a canonical form such that standard algorithms may be applied to estimate the multi-resolution connectivity structure. Therefore, the goal is to form a parameterized equation for coefficients of the field basis functions $\mathbf{x}_t$, which becomes the state vector.

@@

\section{B-Spline MRAIDE in State-Space }
\subsection{A Brief Introduction to B-Spline Wavelets}
B-spline functions are one of the most basic elements for wavelet construction. These functions are composed of piecewise polynomials combined smoothly at some breaking points (knots), which are equally spaced in the case of cardinal B-splines. The degree of smoothness is determined by the order of splines \citep{Goswami1999}, where the higher the order the smoother the resulting B-spline becomes (see Figure~\ref{fig:Figure0}).
\begin{figure}[!t]
\centering
\includegraphics{./Graph/Figure0.pdf}
% \includegraphics{./Graph/Figure0b.pdf}
\caption{{\bf Examples of cardinal B-spline functions}. Piecewise polynomial functions (colored lines) are joined at the break points (black dots). (\textbf{A}) B-spline function of order 2 (linear B-spline). (\textbf{B}) B-spline function of order 8.}
\label{fig:Figure0}
\end{figure}

B-spline wavelet and scaling functions were formulated independently by \citet{Chui1992b,Chui1992,Unser1993}.  The scaling functions are $m$-th order B-splines and the compact support wavelet functions are a linear combination of scaling functions. B-spline wavelet and scaling functions approach optimal time/frequency localization as the order of the spline increases. In fact, for the cubic B-spline scaling and wavelet functions it is already close to the optimal limit for Gaussian functions and they have a better approximation rate than other wavelets with the same number of vanishing moments \citep{Unser1999}.

B-spline wavelet and scaling functions are particularly suited in this framework due to the property of being able to analytically define the convolution and inner product to form MRAIDE components. The following summarizes some important characteristics of B-splines relevant to the proposed modeling framework.

@@

By exploiting \eqref{eq:BsplineConvolution} and \eqref{eq:BsplineInnerProduct}, the integrals in \eqref{eq:Lambdax}, \eqref{eq:Lambdatheta} and \eqref{eq:CovMatrix} can be computed analytically. In constructing $\boldsymbol\Lambda_{x}$, it is important to note that B-spline scaling and wavelet functions possess the following orthogonality properties \citep{Unser1993}: 
\begin{equation}
  \left\langle \varphi_{m;j_1,l_1}(r),\varphi_{m;j_2,l_2}(r)\right\rangle =0  \quad \mathrm{for} \quad j_1\neq j_2
 \label{eq:MRA-PsiPsiOrthogonality} 
 \end{equation}
 \begin{equation}
  \left\langle N_{m;j_1,l_1}(r),\varphi_{m;j_2,l_2}(r)\right\rangle =0  \quad \mathrm{for} \quad j_1\leq j_2.
 \label{eq:MRA-PhiPsiOrthogonality}
 \end{equation}
  \parham{Note from \eqref{eq:MRA-PsiPsiOrthogonality} B-spline wavelets are semi-orthogonal as they are not orthogonal with respect to translation on a given scale.}

@@

To demonstrate the MRAIDE framework we shall consider the parameters to be static and we shall also assume that the relationship between the mean membrane potential and the mean firing rate is predominantly linear. Note that there is a difference between forward and inverse modeling when making these assumptions. For example, by using a linear activation function in the estimator we are not assuming that there is necessarily a linear relationship between the membrane voltage and firing rate, but that the statistics of the neural masses fall within a linear operating region over the estimation time period.

@@

By assuming a linear activation function of the form
\begin{equation}
	f(v_t(\mathbf{r})) = \varsigma v_t(\mathbf{r})
\end{equation}
the state transition equation~\eqref{eq:ApproxDiscreteTimeModel4} simplifies to
\begin{equation}\label{eq:ApproxDiscreteTimeModel_Linear}
	\mathbf{x}_{t+1} = 
	\xi \mathbf{x}_t + 
	\varsigma T_s \mathbf{\Lambda}_{x}^{-1} \int_{\Omega}\boldsymbol\mu\left(\mathbf{r}\right)\int_\Omega { 
	    \boldsymbol\theta^\top\boldsymbol\lambda\left(\mathbf{r}-\mathbf{r}'\right)
	    \boldsymbol\mu^\top\left(\mathbf{r}'\right) 
	\, \mathrm{d}\mathbf{r}'\mathrm{d}\mathbf{r}} \mathbf{x}_t
	+ \mathbf{w}_t.
\end{equation}
Now we can further simplify the system by defining
\begin{align}
	\label{eq:Lambdatheta}
	 \mathbf{\Lambda}_{\theta} &\triangleq \int_{\Omega}\boldsymbol\mu\left(r\right) \int_\Omega { 
		   \boldsymbol\theta^\top\boldsymbol\lambda\left(r-r'\right)
		    \boldsymbol\mu^\top\left(r\right)\ \mathrm{d}r'\mathrm{d}r}
\end{align}
to give the state transition equation
\begin{align}\label{eq:StateEquation}
 \mathbf x_{t+1} &=\mathbf A(\boldsymbol \theta) \mathbf x_t+ \mathbf w_t\\
\label{eq:A_theta}
 \mathbf A(\boldsymbol \theta) &= T_s\varsigma\mathbf{\Lambda}_{x}^{-1}\mathbf{\Lambda}_{\theta}+\xi\mathbf I.
\end{align} 
The problem of joint state and parameter estimation for the state-space model given by \eqref{eq:StateEquation} and \eqref{eq:ReducedObservationEquation} can be formulated as 
\begin{equation}
	{\hat{\mathbf X},\hat{\boldsymbol\theta}}=\arg\max_{\mathbf X,\boldsymbol\theta}p(\mathbf Y;\boldsymbol\theta),
 \end{equation}  
where $\mathbf X$ and $\mathbf Y$ denote the entire sequences of the states and observations respectively, $\boldsymbol\theta$ is the parameter set and $p(\mathbf Y;\boldsymbol\theta)$ denotes the probability density function of the observations parameterized  by $\boldsymbol\theta$, known as the likelihood function. This problem cannot be solved separately as for the state estimation the knowledge of the system's parameters is required and for the parameter estimation the states of the system need to be known. One approach is to iteratively estimate the states and parameters of the system and monitor the convergence of the algorithm. A natural framework to perform this is the well known Expectation Maximization (EM) algorithm \cite{Dempster1977,Shumway2000} to infer both the parameter set and the states from the observations corresponding to the connectivity kernel, the neural field and electrophysiological data in our application. There are two main applications of the EM algorithm:  first is when the observations are incomplete due to a faulty observation process or existing limitations and the second  when the direct maximization of the likelihood function is difficult but can be simplified by assuming the existence of an additional however missing (or hidden) variables \cite{Bilmes1998}, the weights of the field basis functions in this case. The EM algorithm, when used in the latter context \cite{Dewar2009}, yields the maximum likelihood kernel estimate, i.e.
\begin{equation}
	\boldsymbol\theta_{\text{ML}}=\arg\max_{\boldsymbol\theta}~p(\mathbf Y;\boldsymbol\theta),
 \end{equation}   
and the posterior distribution of the field over time.

@@

In summary, to solve the joint state and parameter estimation for linear state-space models one can adapt the EM algorithm, solving the smoothing problem for the E-step and constructing and maximizing the lower bound in the M-step. A schematic representation of the EM algorithm is represented in Figure~\ref{fig:EMBlockDiagram}.  

@@

\caption{{\bf Schematic of the EM algorithm}. The RTS smoother (E-step) and the maximization of the lower bound, $\mathcal Q(\boldsymbol \theta,\boldsymbol\theta')$, (M-step) are shown. The dashed box shows the iterative two step algorithm once it is initialized.}

@@

  As mentioned earlier the E-step for state-space models is equivalent to the smoothing problem. We use the RTS smoother in order to infer the states of the system. The RTS smoother uses the Kalman filter \cite{Kalman1960} in forward pass along withe a backward iteration to calculate its outputs: smoothed states $\hat{\mathbf x}^b_t$ and covariances, $\mathbf P^b_t=\mathrm{cov}(\mathbf{x}_t)$. The RTS  algorithm is given in Table~\ref{alg:MRA-RTS}. Following \cite{Gibsona2005}, an additional backwards recursion is also included (Step~5 of the RTS algorithm)	to calculate cross-covariance matrix, $\mathbf M_t=\mathrm{cov}(\mathbf{x}_{t},\mathbf{x}_{t+1})$, which is also required in forming the lower bound to be maximized in the M-step. Note that the following quantities from the forward iteration (filtering) should be stored when implementing the RTS smoother: predicted and corrected estimates for both state estimates ($\mathbf{x}_t^{f-}$, $\mathbf{x}_t^{f}$) and covariance matrices ($\mathbf P_t^{f-}$,$\mathbf P_t^f$). Also the final value of the Kalman gain, $\mathcal K_T$, should be stored in order to initialize the backward cross-covariance computation. 

@@

\caption{\textbf{Algorithm for the RTS smoother} This table shows the steps in the  Rauch-Tung-Striebel smoother algorithm. The lower bound is maximized after each iteration to update the parameter estimates.}

@@

where the constant term can be ignored as it is independent of the parameters. Under the condition that the state distribution is Gaussian --- ignoring the normalizing term, $1/(2\pi)^{\frac{n_x}{2}}\mid\boldsymbol\Sigma_w\mid^{-\frac{1}{2}}$ --- the conditional distribution $p(\mathbf x_{t+1} | \mathbf x_{t};\boldsymbol\theta)$ can be written as

@@

Rearranging and taking the expectation of the log-likelihood function over all time instants gives the required lower-bound which is to be maximized for the optimal parameter estimates. Note that the expectation distributes over the trace sum. Therefore, 

@@

\subsection{Stopping rule}
The stopping criterion is usually associated with either the change in the parameter estimates or the log-likelihood variation \cite{McLachlan1997}. The stopping rule adopted herein is
\begin{equation}
 \left(\parallel \mathbf{A} \parallel_{F}^{(i)}-\parallel \mathbf{A} \parallel_{F}^{(i-1)}\right)<\epsilon,
 \end{equation}
 where $\epsilon$ is a threshold value and $\parallel \mathbf{A} \parallel_{F}^{(i)}$ and $ \parallel \mathbf{A} \parallel_{F}^{(i-1)}$ are the Frobenius norms  of the successive estimates of $\mathbf{A} $ matrices which is defined by \citet{Meyer2000}
 \begin{equation}
  \parallel \mathbf{A} \parallel_{F}=\sqrt{\sum_{i,j=1}^{n_x}\mid a_{i,j} \mid^2}=\sqrt{\mathrm{tr} (\mathbf A^{\top}\mathbf A)}.
 \end{equation}
\subsection{Computational complexity} 
All $n_x^2$ blocks of $\mathbf U$, $\boldsymbol\Gamma_1$ and $\boldsymbol\Gamma_2$  can be computed as a one-off before the commencement of the EM iterations, which increases the speed of the M-step significantly compared to the implementation in \citet{Dewar2009}. The complexity of each step of the algorithm is summarized in Table~\ref{table:MRA-ComputationalComplexity} where for ease of comparison the computational complexity of the method in \citet{Dewar2009} is also identified. It can be seen from the table that the state dimension is an important factor at each iteration of the estimation algorithm, emphasizing the advantage of the proposed algorithm compared to \citet{Dewar2009}. This also suggests the use of an efficient algorithm to choose the minimum number of basis functions (states of the system) while achieving the desired details of the underlying field.  
\begin {table}
\begin{center}
\scalebox{1}{\begin{tabular}{lllc}
\hline \hline
Variable &Equation&Order&Order obtained from \citet{Dewar2009}\\ \hline\\
$\Xi_0, \Xi_1$&\eqref{eq:MRA-Xi0}, \eqref{eq:MRA-Xi1}&$O(Tn_x^2)$& -\\
$\boldsymbol\upsilon_0, \boldsymbol\upsilon_1$&\eqref{eq:epsilon0},\eqref{eq:epsilon1} &$O(n_x^2n_{\theta})$ &$O(n_x^2n_{\theta})+O(n_x^3)$ \\
$\boldsymbol\Upsilon$&\eqref{eq:Epsilon}&$O(n_x^2n_{\theta}^2)$&$O(n_x^4n_{\theta}^2)$\\
$\mathbf A(\boldsymbol\theta)$&\eqref{eq:A_theta}&$O(n_x^3)$&-\\
$\boldsymbol\Lambda_{\theta}$&\eqref{eq:Lambdatheta}&$O(n_x^2n_{\theta})$&-\\
$\hat{\boldsymbol\theta} $&\eqref{eq:MRA-thetahat}&$O(n_{\theta}^3)$&-\\
\hline \hline
\end{tabular}}
 \caption {{\bf The Computational complexity of the estimation algorithm}. A comparison between the algorithm proposed herein and that of \citet{Dewar2009}.} 
\label{table:MRA-ComputationalComplexity}
\end{center}
\end {table}

@@

Two different experiments were designed to test the framework. In the first one, 200 realizations of 1 second of data were used, where the field disturbance, $e_t$, and the observation noise, $\boldsymbol\varepsilon_t$, were regenerated each run. This way the net shape of the connectivity kernel could be formed by averaging over the parameters estimates from each realization. In the second experiment, a single data set was generated and the field estimations are compared for five models, each accounting for a different level of approximation, i.e. $j=0$ to $j=4$.

For each of the experiments, the estimation was applied to the final 900~ms, allowing the forward model's dynamics to stabilize from the initial conditions. The initial state, parameters and the support of the connectivity kernel were unknown to the estimator. The observation noise was set to $\boldsymbol\Sigma_{\varepsilon}=0.1 \times \mathbf{I}_{n_y}$ and the disturbance covariance function was set to $\gamma(r-r') = \varphi_{3,-2}(r-r')$ (B-spline, $j=3$).  The firing rate slope, $\varsigma$ was set to $0.56~mV^{-1}$ \citep{Wendling2005}. The observation kernel was modeled using a B-spline function, with a width of 0.8~mm at half the maximum amplitude where the distance between adjacent sensors was $0.5$~mm, resulting into $n_y = 161$ observations. The spacing and bandwidth of the sensors allowed the full spatial bandwidth of the field to be observed.  With this number of observations, the estimation problem was well-posed, i.e. the number of states, $n_x < n_y$ at each level of the decomposition. This is because the scaling functions are orthogonal to the wavelets at the same and higher levels of decomposition, and  the wavelets are orthogonal to each other across different scales (see (\ref{eq:MRA-PsiPsiOrthogonality}-\ref{eq:MRA-PhiPsiOrthogonality})).

@@

The actual connectivity kernel and the decomposition is plotted in Figure~\ref{fig:KernelEstimate}(a). No assumptions where made about the shape of the kernel. The reconstructed kernel is in good accordance with the actual kernel, where the actual kernel lies inside the confidence interval. The large standard deviation is due to the high number of parameters need to be estimated. The small error in the estimate is attributed to the MRA of the system used to form the estimator. Figure~\ref{fig:KernelEstimate}(b) and (c) illustrate the kernel scaling and wavelet functions, respectively. A total number of 25 basis functions was used to reconstruct the kernel, comprising 13 and 12 scaling and wavelet functions at $j=1$ accordingly. 
\begin{figure}[!t]
\includegraphics{./Graph/Figure5.pdf}
\includegraphics{./Graph/Figure6.pdf}
\caption{ {\bf Connectivity kernel estimate}. (\textbf{a}) The actual kernel and its components are shown by solid and dashed black lines, respectively. The mean kernel estimates over 200 realizations and confidence interval are shown by the red line and red shaded region ($\pm2$ std.). (\textbf{b}) Kernel scaling and wavelets functions, respectively, at $j=1$ used in the estimation algorithm.}
\label{fig:KernelEstimate}
\end{figure}
